Control of Prosumer Networks

### 2016-02-15 Article.tex
intro modifs

Modernizing the power grids and increasing the share of renewables in the production are two substantial objectives of the energetic transition. Because of parallel progresses in communication, data management, and storage, the upcoming emergence of a power grid "2.0", often called smart grid, appears as a cross-disciplinary challenge of the 21st century \cite{Ramchurn}.

@@

Relying on the ability of the smart grid to obtain, communicate and analyse data, several solutions could  improve the stability of such complex systems. Among these solutions, demand side management and energy storage are two of the most popular ones. In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes. More precisely, we are interested in the "best" locations for the storage equipments such that we are able to maintain stability when  perturbations occur. Since all nodes can switch from generator to load and vice versa, the optimal location is susceptible to change with the prosumers power output.
Our goal is then to find the one that, on average, provide the best results.

Since charging and discharging a battery is costly, we are particularly interested in the energy necessary to control the system when perturbations occur. The notion of performance of a given set of locations for using storage would then be inversely proportional to the energy used. One problem with this definition is that it depends explicitly on the state of the system and the prosumers power distribution. Besides, as the system size increases, the number of possible location sets increases exponentially. This forbid any algorithms using complete enumeration and evaluation of the sets. Moreover, we consider  that the electrical lines have maximum capacities, and storage equipments have both a maximum charge / discharge rate and  bounded capacities. Obviously, only solutions satisfying these constraints will be considered as feasible.

 In this paper, we model the prosumer network as a coupled oscillator network. Inspired by \cite{Filatrella2008}, we use a second order Kuramoto model that can be seen as a simplified version of the underlying power grid dynamics (see Section \ref{sec:The_Model}). We then use control theory \cite{Liu2015} in order to model the impact of the storage equipments on the network (see Section \ref{sec:control_and_submodularity}). Using the relationships between submodular set functions and the Gramian matrix \cite{Summers2014}, we use a modified version of a greedy algorithm \cite{Minoux} in order to find the smallest locations set that minimizes the average control energy while satisfying all the constraints (see section \ref{sec:Control_prosumer}). Finally, we show some results in Section \ref{sec:Results}.

@@

For a direct illustrative application, we chose a small system of 15 nodes connected with an erdos-renyi topology. The power distribution and the capacities of the lines are selected randomly such that the system is at equilibrium. Therefore, the production matches the consumption, no line is overloaded and all nodes are synchronized to the main frequency. We use the greedy algorithm (X) based on the trace of the inverse Gramian as to select the driver set $S$. At a given time $ t_k $, we introduce a small perturbation in the form of a mismatch between the production and consumption : $P_i(t_k) = P_i(t_{k-1}) + \delta P$ for some random node i. Consequently, the frequencies $ \dot{ \theta }_i \forall i $ of the oscillators start to deviate from the synchronized state. At time $ t_k + d $, we apply to the drivers the optimal control of equation (X) such that the system is brought to the synchronized state at time $ t_k + d + T $, where T is the control time. At this time, if the mismatch between production and consumption is still active, we maintain synchrony by using optimal control with $T=1$. Figure 1 shows the evolution of a subset (for clarity) of the nodes' frequencies during the control phase.

@@

The purpose of selecting the drivers according to a Gramian related metric is to minimize, on average, the amount of control energy needed. In other words, if we select the drivers randomly, we expect to need, on average, more energy to control the system. In figure 2, we draw 200 example systems of 15 nodes with random topologies, power distributions, and line capacities. These systems are initially at equilibrium and are perturbed as explained previously. For each system we find the driver set S using algorithm (X) and we draw a random driver set $S'$ such that $S$ and $S'$ have the same size, and such that $S'$ also provide full control of the system. We then compute the control signals and the corresponding control energies $\mathcal{E}_S$ and $\mathcal{E}_{S'}$ for both $S$ and $S'$. Each system can therefore be represented as a point with coordinates $(|S|, \mathcal{E}_S)$. On figure 2, we plot the densities of these points for the random case (left pannel) and the optimized case (right pannel). Regions in dark red contains a lot of points while regions in dark blue have very few points. Figure 2 shows that most of the systems considered tend to be controlled by the drivers of algorithm X with less energy and/or drivers than for the random case.

\begin{figure}
\label{fig:densities}
\includegraphics[scale=.3]{rr}
\caption{test}
\end{figure}

@@

\begin{itemize}
\item Level 1 : full control
\item Level 2 : full control + line capacities constraints + battery constraints (see algorithm X)
\end{itemize}

@@

Figure 4 shows how the number of drivers evolves with the number of clusters for different values of $ p_{in} $ when $ p_{out} $ is fixed ($p_{out} =0.1$). For small values of $ p_{in} \sim p_{out} $ clusters are poorly marked, and the connectivity is low. The number of drivers in these conditions is large and increases slowly when the number of clusters augments. As $ p_{in} $ increases, the clusters becomes more densily connected such that within a cluster less drivers are required in order to control it. As the number of clusters grows, more drivers are required. For large values of $p_{in} $, the behavious is more complex. We see indeed that for $p_{in} = 0.9 $ and $N_{clusters} = 2$, almost $ 94\% $ of the nodes are needed for control, but this quantity first decreases with the addition of a few more clusters ($ 71\% $ for $ N_{clusters}=5 $), before increasing. This behavior can be explained by the fact that controlling a very dense network requires, in our settings, a very large portion of nodes to be drivers. Controlling one cluster alone thus requires to control almost all nodes within this cluster. Nevertheless, when a relatively small (compared to the number of nodes in the graph) number of clusters are interconnected with a few links, nodes in one cluster are able to control nodes in other clusters to which they are connected. As the number of clusters grows, the global connectivity increases rapidely, such that the number of drivers also rises. We show this behavior in more details in figure 5, where the number of drivers is plotted against $ p_{in} $ for $N=200$ and $N_{clusters} \in \{2,4,5\} $. For $N_{clusters} = 2 $ and $ p_{in} = 0.1 $ the number of drivers is large since the graph is poorly connected. When $p_{in} $ increases, $n_D$ decreases until a minimum value $n_D \sim 0.64 $ at $ p_{in} \sim 0.6 $. After this point, $n_D$ augments with $p_{in}$. For $N_{clusters} = 5 $, we do not see this behavior : $ n_D $ decreases continuously as $p_{in} $ increases as expected from the curves of figure 4.

@@

Until now, we have considered all batteries to be equal, i.e they have the same capacity and charge/discharge rate. Although simple, this assumption might not be verified in practice where different types of batteries exist. Making a cost-oriented study where the location os various types of batteries is studied such that the tradeoff between costs and reliability is out of the scope of this paper. Nevertheless, we studied how the number of drivers behaves when the batteries have different capcities. More precisely, we draw the capacities of the batteries from a normal distribution $ \mathcal{N}( \mu_{\lambda}, \sigma_{\lambda} ) $ and keep a simple erdos-renyi topology for the power grid. Note that each node is assigned a capacity regardless of its caracteristics (degree, betweeness, and so on...). In figure 6, we plot $n_D$ as a function of the relative standard deviation of the battery capacity distribution $ \sigma_{\lambda} / \mu_{\lambda} $. We see that $n_D$ increases with the variance of the capacity distribution for all three gramian based metrics considered. Moreover, this rise is abrupt since $n_D$ goes from $ 60 \%$ for $\frac{\sigma_{\lambda}}{\mu_{\lambda}} \sim 0.15 $ to $100 \% $ for $\frac{\sigma_{\lambda}}{\mu_{\lambda}} \sim 0.27 $.

>>>>

Modernizing the power grids and increasing the share of renewables in the production are two substantial objectives of the energetic transition. Because of progresses in communication, data management, and storage, the upcoming emergence of a power grid "2.0", often called smart grid, appears as a cross-disciplinary challenge of the 21st century \cite{Ramchurn}.

@@

Relying on the ability of the smart grid to obtain, communicate and analyse data, several solutions could  improve the stability of such complex systems. Among these solutions, demand side management and energy storage are two of the most popular ones. In this paper, we explore the use of storage in a network of prosumers whose power outputs are susceptible to change due to various external causes.

%More precisely, we are interested in the "best" locations for the storage equipments such that we are able to maintain stability when  perturbations occur. Since all nodes can switch from generator to load and vice versa, the optimal location is susceptible to change with the prosumers power output. Our goal is then to find the one that, on average, provides the best results.

%Since charging and discharging a battery is costly, we are particularly interested in the energy necessary to control the system when perturbations occur. The notion of performance of a given set of locations for using storage would then be inversely proportional to the energy used. One problem with this definition is that it depends explicitly on the state of the system and the prosumers power distribution. Besides, as the system size increases, the number of possible location sets increases exponentially. This forbid any algorithms using complete enumeration and evaluation of the sets. Moreover, we consider  that the electrical lines have maximum capacities, and storage equipments have both a maximum charge / discharge rate and  bounded capacities. Obviously, only solutions satisfying these constraints will be considered as feasible.

We model this network as a coupled oscillator network, where each node is represented as an oscillator with a phase angle and a natural frequency that are impacted by the ones of their neighbors. Since the work of Kuramoto [], networks of coupled oscillators have been used to model various phenomenons such as fireflies, ex2, or power grids. Indeed, it is well-known that synchronization of all rotating elements to a common frequency is necessary within a power grid for stable operation. Inspired by \cite{Filatrella2008}, we use a second order Kuramoto model in order to obtain a simplified dynamics for the power grid (see Section \ref{sec:The_Model}).

We will see that under some conditions, the system synchronizes to a common frequency. However, this synchronized state is suceptible to break if perturbations occur. If, for example, the production of some generator decreases while the load remains the same, the imbalance between production and consumption might drive the system to unstable conditions. These kinds of perturbations are very likely in a scenario with high penetration of DER. Multiple control techniques, with different time scales, exits in order to cope with such events. Among others, it has been argued that storage and vehicle to grid (V2G) technology could be used for frequency regulation. If a sudden drop or raise of the frequency due to an imbalance is detected, some devices are charged or discharged as to restore synchrony. This approach poses a few questions such as the location of the storage devices, or the quantities of power that should be injected or absorbed. Since charging and discharging a battery is costly, we are particularly interested in the energy necessary to restaure synchrony. Moreover, in a prosumer scenario where generators and loads tend not to be fixed, we would like to find a subset of locations for installing storage such that the energy is low on average.

We will see that these storage actions can be interpreted as an additional term in the oscillators dynamics for a subset of nodes. We propose here to use complex network control theory \cite{Liu2015} in order to model the impact of the storage equipments on the network dynamics (see Section \ref{sec:control_and_submodularity}). Placing storage at some node can thus be seen as selecting this node in the driver set. Selecting the drivers that, on average, will control the network with low energy is a hard problem. Nevertheless, recent studies showed the existence of relationships between controllability and submodular set functions \cite{Summers2014}. These functions, that we introduce in more depth in section X, have the interesting property of having diminishing returns as we increase the size of the set considered. Although finding the optimal set is NP-hard, \cite{Minoux} proposed a simple greedy algorithm with a worst case guaranty.

Since the system under study is a power grid which lines and batteries have finite capacities, we have further constraints on the control signals. Obviously, no line can be overloaded while trying to restaure synchrony. Likewise, no battery can inject power when empty, or charge/discharge quicker than some maximum rate. We propose here to incorporate these constraints in the search of the driver set (see section \ref{sec:Control_prosumer}).

This paper is organized as follows, section \ref{sec:The_Model} presents the oscillator model used for simplifying the grid's dynamics. Section 3 introduces submodular functions and their maximization. Section 4 gives basic notions of control theory and links controllability to submodular set functions. In section \ref{sec:Control_prosumer} we develop the constraints relative to the system and the batteries. Finally, in section \ref{sec:Results}, we show some results.

@@

For a direct illustrative application, we chose a small system of 10 nodes connected with an erdos-renyi topology. The power distribution and the capacities of the lines are selected randomly such that the system is at equilibrium. Therefore, the production matches the consumption, no line is overloaded and all nodes are synchronized to the main frequency. We use the greedy algorithm (X) based on the trace of the inverse Gramian as to select the driver set $S$. At a given time $ t_k $, we introduce a small perturbation in the form of a mismatch between the production and consumption : $P_i(t_k) = P_i(t_{k-1}) + \delta P$ for some random node i. Consequently, the frequencies $ \dot{ \theta }_i \forall i $ of the oscillators start to deviate from the synchronized state. At time $ t_k + d $, we apply to the drivers the optimal control of equation (X) such that the system is brought to the synchronized state at time $ t_k + d + T $, where T is the control time. At this time, if the mismatch between production and consumption is still active, we maintain synchrony by using optimal control with $T=1$. Figure 1 shows the evolution of a subset (for clarity) of the nodes' frequencies during the control phase.

@@

The purpose of selecting the drivers according to a Gramian related metric is to minimize, on average, the amount of control energy needed. In other words, if we select the drivers randomly, we expect to need, on average, more energy to control the system. In figure 2, we draw $10^4$ example systems of 50 nodes with random scale-free topologies (Barabasi-Albert model is used), and random power distributions and line capacities.  For each system we select a random number of drivers $N_D \sim \mathcal{U}(1,N_{nodes}) $ and we find two driver sets of size $N_D $. One is chosen randomly and the other is found by greedy optimization of a gramian based metric. For both sets, we draw a random initial state $Y_i$ and a random final state $Y_f$, and we compute the control energy required for driving the system from $Y_i$ to $Y_f$. On figure \ref{fig:densities} we plot each system as a dot with coordinates $(n_D = N_D / N_{nodes}, log(\mathcal{E}))$. The color indicates what method has been used, and the lines show the average accross $n_D$. As visible, both curve decreases as the number of drivers increases, meaning that, as the number of drivers grows, the average control energy tends to decrease. Furthermore, we tend to use, as exected, less energy when the drivers are carefully selected. Note that this difference tends to zero when $n_D$ tend to one, because almost all nodes are then selected, yielding little flexibility for optimization.

%For each system we find the driver set S using algorithm (X) and we draw a random driver set $S'$ such that $S$ and $S'$ have the same size, and such that $S'$ also provide full control of the system. We then compute the control signals and the corresponding control energies $\mathcal{E}_S$ and $\mathcal{E}_{S'}$ for both $S$ and $S'$. Each system can therefore be represented as a point with coordinates $(|S|, \mathcal{E}_S)$. On figure 2, we plot the densities of these points for the random case (left pannel) and the optimized case (right pannel). Regions in dark red contains a lot of points while regions in dark blue have very few points. Figure 2 shows that most of the systems considered tend to be controlled by the drivers of algorithm X with less energy and/or drivers than for the random case.

\begin{figure}
\label{fig:densities}
\includegraphics[scale=.35]{energy}
\caption{Control energy $log(\mathcal{E})$ against size of control set $n_D$ for random and optimized driver sets. Topologies are drawn from a Barabasi-Albert model with $ N_{nodes} = 50 $, power and line capacity distributions are random. For both case, $10^4$ instances are considered and the results are displayed as dots. The lines shows the average for both case. }
\end{figure}

@@

\begin{itemize}
\item Level 1 : full control : the system can be moved from any point to any other point of the state space. We do not consider constraints on the energy which can tend to infinity. 
\item Level 2 : full control + line capacities constraints + battery constraints (see algorithm X). This level is much more demanding since we impose the ability to move the system from any point to any other point, without overloading any line or breaking any battery limits in the process.
\end{itemize}

@@

Figure 4 displays how the number of drivers evolves with the number of clusters for different values of $ p_{in} $ when $ p_{out} $ is fixed ($p_{out} =0.1$). For small values of $ p_{in} \sim p_{out} $ clusters are poorly marked, and the connectivity is low. The number of drivers in these conditions is large and increases slowly when the number of clusters augments. As $ p_{in} $ increases, the clusters becomes more densily connected such that within a cluster less drivers are required in order to control it. As the number of clusters grows, more drivers are required. For large values of $p_{in} $, the behavious is more complex. We see indeed that for $p_{in} = 0.9 $ and $N_{clusters} = 2$, almost $ 94\% $ of the nodes are needed for control, but this quantity first decreases with the addition of a few more clusters ($ 71\% $ for $ N_{clusters}=5 $), before increasing. This behavior can be explained by the fact that controlling a very dense network requires, in our settings, a very large portion of nodes to be drivers. Controlling one cluster alone thus requires to control almost all nodes within this cluster. Nevertheless, when a relatively small (compared to the number of nodes in the graph) number of clusters are interconnected with a few links, nodes in one cluster are able to control nodes in other clusters to which they are connected. As the number of clusters grows, the global connectivity increases rapidely, such that the number of drivers also rises. We show this behavior in more details in figure 5, where the number of drivers is plotted against $ p_{in} $ for $N=200$ and $N_{clusters} \in \{2,4,5\} $. For $N_{clusters} = 2 $ and $ p_{in} = 0.1 $ the number of drivers is large since the graph is poorly connected. When $p_{in} $ increases, $n_D$ decreases until a minimum value $n_D \sim 0.64 $ at $ p_{in} \sim 0.6 $. After this point, $n_D$ augments with $p_{in}$. For $N_{clusters} = 5 $, we do not see this behavior : $ n_D $ decreases continuously as $p_{in} $ increases as expected from the curves of figure 4.

@@

Until now, we have considered all batteries to be equal, i.e they have the same capacity and charge/discharge rate. Although simple, this assumption might not be verified in practice where different types of batteries exist. Optimizing the locations of different types of storage as to optimize the cost / reliability tradeoff is out of the scope of this paper. Nevertheless, we studied how the number of drivers behaves when the batteries have different capacities. More precisely, we draw the capacities of the batteries from a normal distribution $ \mathcal{N}( \mu_{\lambda}, \sigma_{\lambda} ) $ and keep a simple erdos-renyi topology for the power grid. Note that each node is assigned a capacity regardless of its caracteristics (degree, betweeness, and so on...). In figure 6, we plot $n_D$ as a function of the relative standard deviation of the battery capacity distribution $ \sigma_{\lambda} / \mu_{\lambda} $. We see that $n_D$ increases with the variance of the capacity distribution for all three gramian based metrics considered. Moreover, this rise is abrupt since $n_D$ goes from $ 60 \%$ for $\frac{\sigma_{\lambda}}{\mu_{\lambda}} \sim 0.15 $ to $100 \% $ for $\frac{\sigma_{\lambda}}{\mu_{\lambda}} \sim 0.27 $.
